\section{Discussion}

\subsection{Limitations of the Framework}

Our five-level framework, like all classification systems, involves simplifications:

\textbf{Discrete vs. Continuous}: Real systems exist on a continuum. A textbook with sophisticated simulations but no adaptivity might be ``Level 2.5.'' The discrete levels should be understood as reference points rather than rigid categories.

\textbf{Domain Dependence}: The optimal level may vary by subject matter. Procedural skills (mathematics, programming) may benefit more from Level 3 adaptivity than humanities subjects where multiple valid interpretations exist.

\textbf{Cultural Context}: Privacy expectations and regulatory environments vary globally. A system appropriate for one jurisdiction may be problematic in another.

\textbf{Rapid Evolution}: AI capabilities are advancing rapidly. Features currently requiring Level 4 infrastructure may become achievable at Level 2 costs within years.

\subsection{Relationship to Learning Graphs}

Our framework is closely related to the concept of learning graphs---structured representations of concept dependencies that enable navigation through educational content.

Learning graphs are a fundamental component of Level 2 intelligent textbooks. At this level, learning graphs serve as static knowledge structures that:
\begin{itemize}
    \item Define prerequisite relationships between concepts
    \item Enable students to visualize the overall course structure
    \item Support non-linear navigation based on student interests
    \item Provide context for where each concept fits in the broader curriculum
\end{itemize}

The key distinction is that at Level 2, learning graphs inform \textit{student-directed} exploration, while at Level 3 and above, they enable \textit{system-directed} adaptation:
\begin{itemize}
    \item \textbf{Level 2}: Learning graphs help students choose their own path; the system does not track which path was taken
    \item \textbf{Level 3}: Systems use learning graphs combined with individual performance data to sequence content and enforce prerequisites automatically
    \item \textbf{Level 4}: Systems use learning graphs to ground chatbot responses in verified concept relationships
    \item \textbf{Level 5}: Systems use learning graphs as the knowledge backbone for autonomous tutoring
\end{itemize}

The quality of the learning graph significantly impacts effectiveness at all levels. At Level 2, a well-structured graph helps students understand concept relationships; at Level 3+, it determines the quality of adaptive pathways. Poorly structured graphs lead to confusion at Level 2 and suboptimal learning paths at higher levels.

\subsection{The Role of Human Instructors}

Our framework should not be interpreted as advocating for replacement of human instructors. Rather, each level changes the instructor's role:

\begin{itemize}
    \item \textbf{Level 1--2}: Instructor as content curator and discussion facilitator
    \item \textbf{Level 3}: Instructor as learning path supervisor and exception handler
    \item \textbf{Level 4}: Instructor as AI overseer and mentor for complex questions
    \item \textbf{Level 5}: Instructor as learning experience designer and pastoral supporter
\end{itemize}

Even at Level 5, human instructors remain essential for motivation, ethical guidance, and handling situations beyond AI capabilities.

\subsection{Equity Considerations}

There is reason for optimism regarding educational equity. The author is confident that within two years, high-quality intelligent textbooks will be freely available for most popular high school and college-level courses under highly permissive Creative Commons licenses that allow each organization to customize content to match their curriculum. Several intelligent textbooks for high school mathematics and other STEM topics are already available today~\cite{mccreary2025casestudies}. This proliferation of open educational resources at Level 2 could dramatically reduce textbook costs while providing interactive, engaging content to students regardless of institutional resources.

However, higher-level intelligent textbooks risk exacerbating educational inequities:

\begin{itemize}
    \item \textbf{Access}: Level 3+ systems require reliable internet and modern devices, potentially disadvantaging students with limited technology access.
    \item \textbf{Data literacy}: Students must understand data collection to provide meaningful consent, requiring digital literacy education.
    \item \textbf{Algorithmic bias}: Systems trained on data from well-resourced institutions may perform poorly for underrepresented populations.
\end{itemize}

Institutions should assess equity implications before adoption and ensure lower-level alternatives remain available.

\subsection{Future Research Directions}

This framework opens several promising research directions that can advance the field of intelligent textbooks:

\begin{enumerate}
    \item \textbf{Empirical validation of level boundaries}: Research studies can validate whether organizations that adopt the five-level model take more concrete actions to increase the sophistication of their educational materials. We anticipate finding a positive correlation between framework adoption and measurable improvements in textbook quality.

    \item \textbf{Standardized assessment instruments}: Development of rating systems for textbooks, departments, and institutions based on average textbook intelligence levels. These transparent metrics will empower parents and students to evaluate the accountability and quality of educational institutions, driving healthy competition toward better learning resources.

    \item \textbf{Privacy-preserving architectures}: Emerging techniques such as federated learning and differential privacy offer promising paths for Level 3+ systems that deliver personalization benefits while protecting student data. As these technologies mature, the privacy threshold may become easier to cross responsibly.

    \item \textbf{Long-term learning outcome studies}: As low-cost intelligent textbooks gain adoption, longitudinal studies comparing learning outcomes across levels will become feasible. While privacy considerations require careful study design, the growing ecosystem of open educational resources provides unprecedented opportunities for rigorous evaluation.

    \item \textbf{Cross-cultural adaptation}: The regions with the fewest educational resources stand to benefit most from freely available intelligent textbooks. Thoughtful adaptation of this framework to diverse cultural contexts---combined with clear communication that AI augments rather than replaces teachers---can accelerate global adoption and narrow educational gaps worldwide.
\end{enumerate}
